---
title: Bootstrap Confidence Intervals for Standardized Solution in lavaan
author: "Shu Fai Cheung"
date: '2022-09-28'
categories: ["R", "lavaan", "bootstrapping", "confidence-intervals", "semhelpinghands", "standardized"]
bibliography: references.bib
csl: apa.csl
aliases:
  - ../../post/2022-09-28-bootstrap-confidence-intervals-for-standardized-solution-in-lavaan/index.html
---


`lavaan` supports bootstrap confidence
intervals for free and user-defined
parameters. This is useful especially for
parameter estimates that may not be
approximately normally distributed unless
the sample size is very large.

However, it is known, though not well-known
enough in my opinion, that, even if bootstrap
confidence intervals are requested, the
confidence intervals reported in the
standardized solution are not bootstrap
confidence intervals as in tools like
`PROCESS` for standardized effects like
standardized indirect effects, but are
symmetric delta-method
confidence intervals based on the
bootstrap
sampling variance-covariance matrix.

Let's use a sample dataset for illustration:

```{r}
# Create the data
set.seed(860541)
n <- 100
x <- rnorm(n)
m <- .4 * x + rnorm(n, 0, sqrt(1 - .3^2))
y <- .4 * m + rnorm(n, 0, sqrt(1 - .4^2))
dat <- data.frame(x = 10 * x, m = 2 * m, y = 3 * y)
```

We specify a simple regression model:

```{r}
library(lavaan)
mod <-
"
m ~ a * x
y ~ b * m + cp * x
ab := a * b
"
```

... and fit it with bootstrap confidence
intervals (2023-01-28: Code and results
updated for `lavaan` 0.6-13, `iseed` is used
instead of `set.seed()`):

```{r}
#| eval: false
fit <- sem(mod, data = dat, fixed.x = FALSE,
           se = "boot", bootstrap = 2000,
           iseed = 8970)
```

```{r}
#| echo: false
fn <- "fit.rds"
if (file.exists(fn)) {
    fit <- readRDS(fn)
  } else {
    fit <- sem(mod, data = dat, fixed.x = FALSE,
              se = "boot", bootstrap = 2000,
              iseed = 8970)
    saveRDS(fit, fn)
  }
```

Let's focus on the confidence intervals of
the indirect effect:

```{r}
est <- parameterEstimates(fit)
std <- standardizedSolution(fit)
# Unstandardized
est[7, ]
# Standardized
std[7, ]
```

They lead to different conclusions.

As shown below, the confidence interval
of the unstandardized indirect effect
is percentile confidence interval that
is asymmetric, as expected:

```{r}
est[7, c("ci.lower", "ci.upper")] - est[7, "est"]
```

However, the confidence interval of the
standardized indirect effect is symmetric:

```{r}
std[7, c("ci.lower", "ci.upper")] - std[7, "est.std"]
```

```{r}
#| echo: false
#| eval: false
fit_boot <- lavInspect(fit, "boot")
boot_ab <- fit_boot[, "a"] * fit_boot[, "b"]
hist(boot_ab)
qqnorm(boot_ab); qqline(boot_ab)
```

This behavior has been discussed
in the [Google group for`lavaan`](https://groups.google.com/g/lavaan) and
so is known, but not "well-known" because
I met many users who were not aware of this,
especially when they use bootstrapping to
get the confidence intervals for indirect
effects but found that the confidence
intervals of unstandardized and
standardized indirect effect led to different
conclusions, as in the example above.

A solution already exists in `lavaan`.
Users can use
`bootstrapLavaan()` and get the bootstrap
confidence intervals for many results,
including
the output of standardized solution.

We first define a function to extract
the standardized indirect effect:

```{r}
fct <- function(fit) {
    lavaan::standardizedSolution(fit)[7, "est.std"]
  }
```

We then update the fit object to disable
standard error because we only need the
point estimates and then call
`bootstrapLavaan()`:

```{r}
#| eval: false
fit0 <- update(fit, se = "none")
fit_boot <- bootstrapLavaan(fit0, R = 2000, FUN = fct, iseed = 8970)
```

```{r}
#| echo: false
fn <- "fit_boot.rds"
if (file.exists(fn)) {
    fit_boot <- readRDS(fn)
  } else {
    fit0 <- update(fit, se = "none")
    fit_boot <- bootstrapLavaan(fit0, R = 2000, FUN = fct,
                                iseed = 8970)
    saveRDS(fit_boot, fn)
  }
```

The percentile confidence interval
can then be formed by `quantile()`.

(Note that `lavaan()` does not use `quantile()` but
use the approach by `boot.ci()`. The resulting
interval can be slightly different from that by `quantile()`.)

```{r}
quantile(fit_boot[, 1], c(.025, .975))
```

However, this is inconvenient because
we need to write
custom function, and
bootstrapping was done twice unless
we store both the unstandardized and
standardized solutions in the custom
function used when calling
`bootstrapLavaan()`.

I wrote the function
[`standardizedSolution_boot_ci()`](https://sfcheung.github.io/semhelpinghands/reference/standardizedSolution_boot_ci.html), available in
the package [`semhelpinghands`](https://sfcheung.github.io/semhelpinghands/), for this
particular
case that I sometimes encounter:

- A model is already fitted with
 `se = "boot"` and so bootstrap confidence
 intervals are already available for the
 unstandardized estimates.

- I want to get the bootstrap
confidence intervals for the
standardized solution *without doing the bootstrapping again*.

This would be useful to
me because some of my projects involve large
samples with missing data. and bootstrapping
takes appreciable time even with
parallelization.

This is how to use this function:

```{r}
library(semhelpinghands)
std_boot <- standardizedSolution_boot_ci(fit)
# -c(9, 10) is used to remove the delta-method CIs from
# the printout
std_boot[, -c(9, 10)]
```

The `boot.ci` intervals are "true"
bootstrap confidence intervals, formed
from the bootstrap estimates. The
bootstrap confidence interval for
the standardized indirect effect
(`r paste0("[", paste(round(std_boot[7, c(11, 12)], 3), collapse = ", "), "]")`)
and that for the unstandardized
indirect effect (`r paste0("[", paste(round(est[7, c(9, 10)], 3), collapse = ", "), "]")`)
now lead to the same conclusion.

`standardizedSolution_boot_ci()` works
like `standardizedSolution()`,
but extracts the stored bootstrap estimates,
get the standardized solution from each
set of estimates, and use them to form
the bootstrap confidence intervals for
the standardized solution.

By default, the bootstrap standardized
solution is also stored in the attribute
`boot_est_std`. They can be extracted
to examine the distribution. For example,
the bootstrap standardized indirect effects
can be extracted and plotted:

```{r}
std_boot_est <- attr(std_boot, "boot_est_std")
std_indirect_boot_est <- std_boot_est[, 7]
hist(std_indirect_boot_est)
qqnorm(std_indirect_boot_est)
qqline(std_indirect_boot_est)
```

This function is simple to use, at least for
me. No need to write custom function,
and no need to do bootstrapping twice.
In most cases, I don't even need to
specify any additional arguments.

More about this function can be found
in the [vignette](https://sfcheung.github.io/semhelpinghands/articles/standardizedSolution_boot_ci.html) for
`standardizedSolution_boot_ci()`.

If any bug in `standardizedSolution_boot_ci()`
was found, I would appreciate submitting
it as a [GitHub issue](https://github.com/sfcheung/semhelpinghands/issues).  {{< emoji ":smile:" >}}