---
title: Plotting Moderation Effects With ggplot2
author: "Shu Fai Cheung"
date: '2022-09-11'
slug: []
categories: ["R", "moderation"]
tags: ["R", "plot", "moderation", "ggplot2"]
output_format:
  html_document:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

There are some R packages that help
researchers to plot moderation (interaction)
effects: The linear relations between
`x` (independent variable /
predictor) and `y` (dependent variable /
outcome variable) for
two or more levels of `w` (moderator). For example, I
have been using [visreg](https://pbreheny.github.io/visreg/)
for multiple regression models fitted by `lm()`
for a long time. It is simple to use and supports
both base R graphics and `ggplot2`.
[stdmod](https://sfcheung.github.io/stdmod/index.html),
which I maintained, also has the function
[plotmod](https://sfcheung.github.io/stdmod/articles/plotmod.html) for plotting simple effects in moderated regression.
For structural equation modelling, [semTools](https://github.com/simsem/semTools/wiki)
can be used to plot interaction for latent
variables using [plotProbe()](https://rdrr.io/cran/semTools/man/plotProbe.html).
`plotProbe()` can also be used to on observed
variables using this [workaround](https://groups.google.com/g/lavaan/c/h_LHQUFbx0k/m/jmQwRJEqBAAJ).

However, there may be case in which all
we need is just two or more lines, and
all we have are the simple effects: Two
or more sets of intercepts of slopes.

This is how to plot the simple effect:

Suppose we want to plot the simple
effects of `x`  on `y` conditional on `w`
(the moderator).
From the output of some functions, we
have the slopes and intercepts
when `w` is "Low" or "High":

`w` is "Low": slope = 1, intercept = 2

`w` is "High": slope = 2, intercept = 3

Write a simple function to compute the points

```{r}
# Simple regression model
xyline <- function(x, a, b) {a + b * x}
```

Set the range for `x`:

```{r}
# Range of x
x <- c(0, 10)
```

Compute the predicted values of `y` at
the lower and upper limit of the range
of `x`, for each level of `w`, when
all other predictors of `y` in the model,
if any, are equal to zero:

```{r}
# Generate the two points when moderator = "Low"
dat0 <- data.frame(Moderator = "Low",
                   x = x,
                   y = xyline(x, a = 0, b = 0.5))
# Generate the two points when moderator = "High"
dat1 <- data.frame(Moderator = "High",
                   x = x,
                   y = xyline(x, a = 2, b = 1.0))
```

Combine the datasets:

```{r}
dat <- rbind(dat0, dat1)
dat
```

Draw the lines using `ggplot2`:

```{r plot_original, tidy = FALSE}
library(ggplot2)
p <- ggplot(dat, aes(x = x, y = y, color = Moderator)) +
            geom_line() +
            scale_color_manual(values = c("Low" = "blue", "High" = "red"))
p
```

This plot can then be modified as necessary:

```{r plot_edited, tidy = FALSE}
p2 <- p + xlab("Independent Variable") +
          ylab("Dependent Variable")
p2
```

This solution can be used for multiple
regression or structural equation modelling.

Though not as elegant as using
packages devoted to plotting moderation
effects, this solution may be good
enough for some simple scenarios. I
believe it can be further improved. However,
if we want more, maybe it is better to
use packages like `semTools` and
`visreg`.

P.S.: This post is based on a suggestion I
posted to the Google Group for `lavaan`.